---
title: "My Code and Outputs"
output:
  html_document:
    toc: true
    toc_float: true
    collapsed: false
    number_sections: false
    toc_depth: 1
    #code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message=FALSE,warning=FALSE, cache=TRUE)
```
Hi! Welcome to where my work is :) 
You can scroll through chronologically, or skip to particular sections you are interested in from the navigation bar on the left - please give it a while to load then mouse-over it. 

# 1. Loading and joining data
Set working directory first


Loading required libraries
```{r}
library(sf)
library(rgdal)
library(raster)
library(mapview)
library(spatstat)
library(sp)
library(rgeos)
library(maptools)
library(GISTools)
library(tmap)
library(sf)
library(geojson)
library(geojsonio)
library(tmaptools)
```

Opening shapefile, taken from ukdataservice.ac.uk, giving a map of Greater Manchester by 2011 English Census Merged Wards (hence data for basemap shapefile is by ward level).
```{r}
require(rgdal)
require(ggplot2)
shp <- readOGR(dsn ="BoundaryData", layer="england_cmwd_2011", stringsAsFactors = F)
plot(shp)
```
```{r}
#checking Coordinate Reference System of the shapefile, which was originally NA
summary(shp)
crs(shp)
```

Specifying CRS of the greater manchester area shapefile as EPSG 4326 for latitude/longiture
```{r}
Manchester <- spTransform(shp, CRS("+init=epsg:4326"))
st_crs(Manchester)
```
Plot Manchester basemap

```{r}
plot(Manchester)
```

Extract just the city of Manchester itself rather than greater Manchester, as the area would otherwise be too big for analysis and Ripley's K. 
```{r}
#extract the city
cityManchester <- subset(Manchester, name %in% c("Didsbury West","Fallowfield","Gorton North","Gorton South","Harpurhey","Higher Blackley","Hulme","Levenshulme","Longsight","Miles Platting and Newton Heath","Moss Side","Moston","Northenden","Old Moat","Rusholme","Sharston","Whalley Range","Withington","Woodhouse Park","Ancoats and Clayton","Ardwick","Baguley","Bradford","Brooklands","Burnage","Charlestown","Cheetham","Chorlton","Chorlton Park","City Centre","Crumpsall","Didsbury East"))

#Check to see that the correct borough has been pulled out
tm_shape(cityManchester) +
  tm_polygons(col = NA, alpha = 0.5)
```




Read in police station data for greater manchester area, from the geoJSON file obtained using overpass-turbo.eu API and running a query for police stations in the Greater Manchester Area from OpenStreetMap. 

Keeping only the points, as readOGR cannot read both points and polygons at the same time, and the points represent police stations.
```{r}
pol <- rgdal::readOGR("C:/Users/Tommy/Desktop/GIS_proj/BoundaryData/polstn.geojson",require_geomType="wkbPoint")
```

Plotting the police station points
```{r}
plot(pol)
summary(pol)
```

Checking the CRS of the police station points - it is already EPSG4326!
```{r}
st_crs(pol)
```
Removing duplicates from police station points

```{r}
polstns <- remove.duplicates(pol)
```

Plotting the police station points on top of the greater manchester area map using mapview
```{r}
mapview::mapview(Manchester) + mapview::mapview(polstns, color = "white", col.regions = "black",legend=FALSE)
```

Basemap of Manchester is a SpatialPolygonsDataFrame
Police station points is a SpatialPointsDataFrame
```{r}
class(cityManchester)
class(polstns)
```

Plot the police stations in the Greater Manchester area using tmap 
```{r}
tmap_mode("view")
tm_shape(Manchester) +
  tm_polygons(col = NA, alpha = 0.5) +
tm_shape(polstns) +
  tm_dots(col = "blue")
```
Plot the police stations in the Manchester city area using tmap 
```{r}
tmap_mode("view")
tm_shape(cityManchester) +
  tm_polygons(col = NA, alpha = 0.5) +
tm_shape(polstns) +
  tm_dots(col = "blue")
```

Filtering for police stations which are within the Manchester city boundary
```{r}
proj4string(cityManchester) <- CRS("+init=epsg:4326")
proj4string(polstns) <- CRS("+init=epsg:4326")

polsub <- polstns[cityManchester,]
#checking to see that they've been removed
tmap_mode("view")
tm_shape(cityManchester) +
  tm_polygons(col = NA, alpha = 0.5) +
tm_shape(polsub) +
  tm_dots(col = "blue")
```

Reading in the dataset for population density 

```{r}

library(readr)
popdens <- read_csv("popdens.csv")
summary(popdens)
```
Joining first dataset of population density with the Manchester basemap 
```{r}
require(sp)
?sp::merge
manchester_popdens <- merge(cityManchester, popdens, by.x = "name", by.y = "geography",duplicateGeoms = TRUE)
```

Plotting map to test if join has succeeded
```{r}
tmap_mode("view")
tm_shape(manchester_popdens) +
  tm_polygons(col = NA, alpha = 0.5) +
tm_shape(polsub) +
  tm_dots(col = "blue")
```

Reading in the dataset for deprivation - the average number (out of 4 - employment, education, health and disability and household overcrowding) of dimensions a household was deprived in was used
```{r}
depr <- read_csv("deprivation.csv")
summary(depr)
```

Joining second dataset of deprivation with the Manchester basemap 
```{r}
manchester_2 <- merge(manchester_popdens, depr, by.x = "name", by.y = "geography",duplicateGeoms = TRUE)
```

Plotting map to check if second join has succeeded
```{r}
tmap_mode("view")
tm_shape(manchester_2) +
  tm_polygons(col = NA, alpha = 0.5) +
tm_shape(polsub) +
  tm_dots(col = "blue")
```
Reading in the dataset for SES - the average social grade by ward was taken, by using a numeric scale of AB = 1, C1 = 2, C2 = 3, DE = 4, so lower numbers would mean a higher social grade.
```{r}
ses <- read_csv("SES.csv")
summary(ses)
```
Joining third dataset of SES with the Manchester basemap 
```{r}
manchester_merged <- merge(manchester_2, ses, by.x = "name", by.y = "geography",duplicateGeoms = TRUE)
```

Plotting map to check if third join has succeeded
```{r}
tmap_mode("view")
tm_shape(manchester_merged) +
  tm_polygons(col = NA, alpha = 0.5) +
tm_shape(polsub) +
  tm_dots(col = "blue")
```

manchester_merged is the map of Manchester city with merged data on population density, social grade and SES.

# 2. Plotting/performing KDE for crime data in Manchester city

Next step would be to map the crime datapoints onto the Manchester basemap. Crime data for June 2019 for the Greater Manchester area was used. This was taken from https://data.police.uk/data/ 




```{r}
crime <- read_csv("manchester_crime.csv")
crime_locations <- st_as_sf(crime, coords = c("Longitude","Latitude"), crs = 4326)
class(crime)
require(sf)
crime_sf <- st_as_sf(x = crime, 
                        coords = c("Longitude", "Latitude"),
                        crs = "+init=epsg:4326")
# simple plot
plot(crime_sf)

crime_spdf <- as(crime_sf, "Spatial")


manchester_crime_test<-ggplot(crime, aes(x=Longitude,y=Latitude))+geom_point()+coord_equal()
manchester_crime_test

```
Plotting a 2D KDE of crime occurences in Manchester using the longitude and latitude data in the crime dataset
```{r}
manchester_crime_test<-ggplot(crime, aes(x=Longitude,y=Latitude))+stat_bin2d(bins=30)
manchester_crime_test
```
Plotting a continuous distribution instead
```{r}
manchester_crime_test+stat_density2d(aes(fill = ..level..), geom="polygon")
```

Plotting the crime datapoints to test
```{r}
plot(crime_locations)
```

Examining metadata of the crime datapoints
```{r}
summary(crime_locations)
```

Checking CRS of crime datapoints
```{r}
crs(crime_locations)
```
Checking type of data of the crimem locations - sf, tbl_df, tbl, data.frame
```{r}
class(crime_locations)
```
Ensuring that crime_locations has EPSG 4326 as CRS
```{r}
crime_locations_sp <- as(crime_locations, 'Spatial')
crime_locations_sp <- spTransform(crime_locations_sp, CRS("+init=epsg:4326"))
```
Checking new CRS of crime datapoints and data type, removing duplicates
```{r}
crs(crime_locations)

```


```{r}
class(crime_locations)
```

Plotting crime data points on Manchester basemap
```{r}
mapview::mapview(manchester_merged) + mapview::mapview(crime_locations, color = "white", col.regions = "black",legend=FALSE, title = "Crime occurences in Manchester city")
```

# 3. Descriptive statistics for data
```{r}
library(tidyverse)
library(downloader)
library(rgdal)
library(sf)
library(ggplot2)
library(reshape2)
library(plotly)
library(highcharter)
```

Histograms for each of the scale variables - population densities, deprivation, and SES
Starting with popdens
```{r}
# Histogram with mean line (red) and median line (blue) and density plot
ggplot(popdens, aes(x=popdens)) + 
 geom_histogram(aes(y=..density..), colour="black", fill="white", binwidth=10)+
 geom_density(alpha=.2, fill="#FF6666") + geom_vline(aes(xintercept=mean(popdens)),
            color="red", linetype="dashed", size=1) + geom_vline(aes(xintercept=median(popdens)),
            color="blue", linetype="dashed", size=1) 

# Descriptive statistics
print(max(popdens$popdens))
print(min(popdens$popdens))
print(median(popdens$popdens))
print(sd(popdens$popdens))
print(IQR(popdens$popdens))
print(quantile(popdens$popdens, c(.25, .75)))

```
Plotting histogram for deprivation next
```{r}
# Histogram with mean line (red) and median line (blue) and density plot
ggplot(depr, aes(x=deprivation)) + 
 geom_histogram(aes(y=..density..), colour="black", fill="white", binwidth= 0.1)+
 geom_density(alpha=.2, fill="#FF6666") + geom_vline(aes(xintercept=mean(deprivation)),
            color="red", linetype="dashed", size=1) + geom_vline(aes(xintercept=median(deprivation)),
            color="blue", linetype="dashed", size=1) 

# Descriptive statistics
print(max(depr$deprivation))
print(min(depr$deprivation))
print(median(depr$deprivation))
print(sd(depr$deprivation))
print(IQR(depr$deprivation))
print(quantile(depr$deprivation, c(.25, .75)))
```
Plotting histogram for social grade next
```{r}
# Histogram with mean line (red) and median line (blue) and density plot
ggplot(ses, aes(x=socialgrade)) + 
 geom_histogram(aes(y=..density..), colour="black", fill="white", binwidth= 0.1)+
 geom_density(alpha=.2, fill="#FF6666") + geom_vline(aes(xintercept=mean(socialgrade)),
            color="red", linetype="dashed", size=1) + geom_vline(aes(xintercept=median(socialgrade)),
            color="blue", linetype="dashed", size=1) 

# Descriptive statistics
print(max(ses$socialgrade))
print(min(ses$socialgrade))
print(median(ses$socialgrade))
print(sd(ses$socialgrade))
print(IQR(ses$socialgrade))
print(quantile(ses$socialgrade, c(.25, .75)))
```
Plotting histogram for crimes recorded per ward next
```{r}
#reading in data for crimes recorded per ward (taken from later part of the code after performing count)
library(readr)
crime_ward <- read_csv("crime_ward.csv")
summary(crime_ward)
crime_ward$lnCrime <- log1p(crime_ward$crimes)
# Histogram with mean line (red) and median line (blue) and density plot
ggplot(crime_ward, aes(x=lnCrime)) + 
 geom_histogram(aes(y=..density..), colour="black", fill="white", binwidth= 0.05)+
 geom_density(alpha=.2, fill="#FF6666") + geom_vline(aes(xintercept=mean(lnCrime)),
            color="red", linetype="dashed", size=1) + geom_vline(aes(xintercept=median(lnCrime)),
            color="blue", linetype="dashed", size=1) 

# Descriptive statistics
print(max(crime_ward$crimes))
print(min(crime_ward$crimes))
print(median(crime_ward$crimes))
print(sd(crime_ward$crimes))
print(IQR(crime_ward$crimes))
print(quantile(crime_ward$crimes, c(.25, .75)))
```

Converting both cityManchester and crime_locations to data.frames to run the KDE, since it requires numeric input
```{r}
df_cityManchester <- as.data.frame(cityManchester)
df_crime_locations <- as.data.frame(crime_locations)
```
Create two separate fields for latitude and longitude from the manchester spatialpolygonsdataframe 
```{r}                                                                                                                                                                
polys = attr(cityManchester,'polygons')
npolys = length(polys)
for (i in 1:npolys){
  poly = polys[[i]]
  polys2 = attr(poly,'Polygons')
  npolys2 = length(polys2)
  for (j in 1:npolys2){
     #do stuff with these values
     coords = coordinates(polys2[[j]])
     
  }
}
coords_df <- as.data.frame(coords)
```
Plotting out crime occurences in Manchestaer city
```{r}
ggplot() + geom_polygon(data = coords_df, aes(x = V1, y = V2), fill = "grey75") +
  geom_point(data = crime, aes(x = Longitude, y = Latitude),
             col = "dodger blue", alpha = .5, size = 1.5) +
  coord_equal() +
  ggtitle("Crime occurences in Manchester")
  
```


Counting number of crime occurence points per polygon
```{r}
class(manchester_merged)
class(crime_spdf)
proj4string(manchester_merged)

crime_spdf <- spTransform(crime_spdf, CRS("+init=epsg:4326"))
proj4string(crime_spdf)
plot(manchester_merged)
plot(crime_spdf, col="red" , add=TRUE)
res <- over(crime_spdf, manchester_merged)
table(res$name)
```
After plotting number of crimes per ward on a .csv file, time to merge with the manchester_merged basemap.
Reading in the dataset for crimes per ward using code chunk directly above
```{r}
summary(crime_ward)
```
Joining crimes per ward dataset with the Manchester basemap 
```{r}
manchester_merged_final <- merge(x= manchester_merged, y=crime_ward, by.x = "name", by.y = "geography")
```

Plotting map to check if join has succeeded. 
```{r}
library(tmap)
tmap_mode("view")
tm_shape(manchester_merged_final) +
  tm_polygons(col = NA, alpha = 0.5) +
tm_shape(polsub) +
  tm_dots(col = "blue")
```
All data (popdens, social grade, SES, crime) on ward level loaded, locations of 4 police stations in Manchester city also loaded. 

# 4. Choropleth maps by variables across Manchester

Making a choropleth map by crimes
```{r}
library(ggplot2)
library(RColorBrewer)
library(classInt)
names(manchester_merged_final)
#spplot(manchester_merged_final, "crimes")
# quantile breaks
breaks_qt <- classIntervals(manchester_merged_final$crimes, n = 7, style = "quantile")
br <- breaks_qt$brks 
offs <- 0.0000001 
br[1] <- br[1] - offs 
br[length(br)] <- br[length(br)] + offs 
# categories for choropleth map
manchester_merged_final$crimes_bracket <- cut(manchester_merged_final$crimes, br)
# plot
#my.palette <- brewer.pal(n = 7, name = "OrRd")
#spplot(manchester_merged_final, "crimes_bracket", col.regions=my.palette, main = "Manchester City Recorded Crimes in June 2019 by Ward")

#Using tmap
library(tmap)
tm_shape(manchester_merged_final) +
  tm_polygons("crimes", 
              style="quantile", 
              title="Manchester city \nrecorded crimes \nby ward in June 2019")
tmap_mode("view")
```
Making a choropleth map by popdens
```{r}
# quantile breaks
breaks_qt <- classIntervals(manchester_merged_final$popdens, n = 7, style = "quantile")
br <- breaks_qt$brks 
offs <- 0.0000001 
br[1] <- br[1] - offs 
br[length(br)] <- br[length(br)] + offs 

# categories for choropleth map
manchester_merged_final$popdens_bracket <- cut(manchester_merged_final$popdens, br)
# plot
#spplot(manchester_merged_final, "popdens_bracket", col.regions=my.palette, main = "Manchester City Population Density by Ward")

#Using tmap
library(tmap)
tm_shape(manchester_merged_final) +
  tm_polygons("popdens", 
              style="quantile", 
              title="Manchester city \nPopulation Density \nby ward")
tmap_mode("view")
```

Making a choropleth map by deprivation
```{r}
#spplot(manchester_merged_final, "deprivation")
breaks_qt <- classIntervals(manchester_merged_final$deprivation, n = 7, style = "quantile")
br <- breaks_qt$brks 
offs <- 0.0000001 
br[1] <- br[1] - offs 
br[length(br)] <- br[length(br)] + offs 
# categories for choropleth map
manchester_merged_final$deprivation_bracket <- cut(manchester_merged_final$deprivation, br)
# plot

#spplot(manchester_merged_final, "deprivation_bracket", col.regions=my.palette, main = "Manchester City deprivation by Ward")
#Using tmap
library(tmap)
tm_shape(manchester_merged_final) +
  tm_polygons("deprivation", 
              style="quantile", 
              title="Manchester city \ndeprivation \nby ward")
tmap_mode("view")
```

Making a choropleth map by social grade
```{r}
#spplot(manchester_merged_final, "socialgrade")
breaks_qt <- classIntervals(manchester_merged_final$socialgrade, n = 7, style = "quantile")
br <- breaks_qt$brks 
offs <- 0.0000001 
br[1] <- br[1] - offs 
br[length(br)] <- br[length(br)] + offs 
# categories for choropleth map
manchester_merged_final$socialgrade_bracket <- cut(manchester_merged_final$socialgrade, br)
# plot
#spplot(manchester_merged_final, "socialgrade_bracket", col.regions=my.palette, main = "Manchester City social grade by Ward")
#Using tmap
library(tmap)
tm_shape(manchester_merged_final) +
  tm_polygons("socialgrade", 
              style="quantile", 
              title="Manchester city \nsocial grade \nby ward")
tmap_mode("view")
```


# 5. Zooming in on Bradford (single ward)
Next, point pattern analysis will be conducted, to confirm the visual suspicion that there is spatial clustering of crimes within Manchester city. 
The entire Manchester city was too large to conduct analysis on, and hence this particular study zoomed in on a particular ward - Bradford. Bradford was picked as it has two police stations, so it would make an interesting study to see if the clusters are near the police stations.

We have to create an observation window for spatstat to carry out analysis within – set to the extent of the Bradford ward.

Thus, pull out the Bradford ward.
```{r}
#extract the Bradford ward
bradford <- manchester_merged_final[manchester_merged_final@data$name=="Bradford",]
#Checking to see if it has been pulled out successfully
tm_shape(bradford) +
  tm_polygons(col = NA, alpha = 0.5)
#sf object:
bradfordSF <- st_as_sf(manchester_merged_final)
bradfordSF <- bradfordSF[bradfordSF$name=="Bradford",]
```



We need to clip the crime occurence datapoints so that we have a subset of just those that fall within the Bradford ward. 
```{r}
#clip the data to our single ward
library(maptools)
crime_respdf <- readShapePoints("crime_exported.shp", proj4string = CRS("+init=epsg:4326 +proj=longlat +datum=WGS84 +no_defs +ellps=WGS84 +towgs84=0,0,0"), verbose = FALSE, repair=FALSE)


polsub_bradford <- polsub[bradford,]
```

Check that it's worked - blue dots are crime occurrences and red dots are police stations in Bradford

```{r}

crime_bradford <- crime_respdf[bradford,]
crs(crime_bradford)
tmap_mode("view")+tm_shape(bradford) +tm_polygons(col = NA, alpha = 0.5) +tm_shape(crime_respdf[bradford,]) +  tm_dots(col = "blue") + tm_shape(polsub_bradford) +tm_dots(col = "red")
```


# 6. KDE for crimes in Bradford
Create an observation window for spatstat to carry out analysis
```{r}
library(spatstat)
library(sp)
library(rgeos)
library(maptools)
library(GISTools)
library(tmap)
library(sf)
library(geojson)
library(geojsonio)
library(tmaptools)
```

```{r}
BNG = "+init=epsg:27700"
bradfordBNG <- spTransform(bradford,BNG)
window <- as.owin(bradfordBNG)
plot(window)
crime_bradfordBNG <-spTransform(crime_bradford,BNG) 
```
For point pattern analysis, have to create a point pattern (ppp) object
```{r}

crime_bradford.ppp <- ppp(x=crime_bradfordBNG@coords[,1],y=crime_bradfordBNG@coords[,2],window=window)
```
Checking the ppp object
```{r}
plot(crime_bradford.ppp,pch=16,cex=0.5, main="Crime occurrences in Manchester city, Bradford ward, June 2019")
class(crime_bradford.ppp)
summary(crime_bradford.ppp)
plot(crime_bradford)
```

Plotting KDE for crime occurences in Bradford ward
```{r}
ds <- density(crime_bradford.ppp)
class(ds)
plot(ds, main = "Crime density in Bradford ward, Manchester city, June 2019")
```

Doing the same with contours
```{r}
K1 <- density(crime_bradford.ppp) # Using the default bandwidth
plot(K1, main=NULL, las=1)
contour(K1, add=TRUE)
```

Changing bandwidth
```{r}
K2 <- density(crime_bradford.ppp, sigma=50) # Using a 50km bandwidth
plot(K2, main=NULL, las=1)
contour(K2, add=TRUE)
```

# 7. Testing for CSR in Bradford
Testing for spatial randomness of crime occurrences in Bradford ward, Manchester city

## 7a. Ripley's K
```{r}
K <- Kest(crime_bradford.ppp, correction="border")
plot(K)
```

Theoretical model of CSR, compare actual results to that. 

## 7b. Kolmogorov-Smirnoff test for CSR
With point data, we specify a real function T(x,y) defined at all locations x,y in our sampling window. We evaluate this function at each of the data points and compare the empirical values of T with the predicted distribution of T under the CSR assumptions. 
```{r}
ks <- cdf.test(crime_bradford.ppp, "x")
plot(ks)
pval <- ks$p.value
pval
```
Density function as a covariate to estimate overall spatial randomness
```{r}
ds <- density(crime_bradford.ppp)
k <- cdf.test(crime_bradford.ppp, ds)
plot(k)
```

## 7c. G function test for CSR
The G function measures the distribution of distances from an arbitrary event to its nearest event (i.e. uses nearest neighbor distances). By plotting our empirically estimated G function against our theoretical expectation if the process were CSR, we can assess the extent to which the empirical distribution of our data is different from the theoretical CSR expectation.
```{r}
gtest <- Gest(crime_bradford.ppp)
gtest
plot(gtest)
```
## 7d. F test for CSR
```{r}
ftest <- Fest(crime_bradford.ppp)
ftest
plot(ftest)
```

# 8. DBSCAN to determine cluster locations 
Carrying out DBSCAN to determine where the clusters are, after having established that CSR does not hold, using the tests above.
```{r}
library(raster)
library(fpc)
library(plyr)
library(OpenStreetMap)
```
Checking CRS of the bradford spatial polygon
```{r}
crs(bradford)
```

```{r}
#first extract the points from the spatial points data frame
crime_bradford_points <- data.frame(crime_bradford@coords[,1:2])
#now run the dbscan analysis
db <- fpc::dbscan(crime_bradford_points, eps = 0.003, MinPts = 30)
#now plot the results
plot(db, crime_bradford_points, main = "DBSCAN Output", frame = F)
plot(bradford, add=T)
```

Can use kNNdistplot() from the dbscan package to find a suitable eps value based on the Ripley's K plot. Above 300metres sees a sharp spike, so 0.003 would be good.
```{r}
library(dbscan)
dbscan::kNNdistplot(crime_bradford_points, k =  30)
```

Using ggplot2 to plot a nicer map
```{r}
library(ggplot2)
```

Calling cluster db object
```{r}
db
db$cluster
```

Adding this information back into crime dataframe
```{r}
crime_bradford_points$cluster <- db$cluster
```
Creating convex hull polygons to wrap around the points in the clusters
```{r}
chulls <- ddply(crime_bradford_points, .(cluster), 
                function(df) df[chull(df$coords.x1, df$coords.x2), ])
```
Can drop 0 from the dataframe, as they are not in clusters
```{r}
chulls <- subset(chulls, cluster>=1)
```
Creating a ggplot object from clusters 1 and 2
```{r}
dbplot <- ggplot(data=crime_bradford_points, 
                 aes(coords.x1,coords.x2, colour=cluster, fill=cluster)) 
#add the points in
dbplot <- dbplot + geom_point()
#now the convex hulls
#convert spatialpointsdataframe to a dataframe as ggplot cannot handle it
df_polsub_bradford <- data.frame(polsub_bradford)
dbplot <- dbplot + geom_polygon(data = chulls, 
                                aes(coords.x1,coords.x2, group=cluster), 
                                alpha = 0.5) + geom_point(data=df_polsub_bradford,aes(x=coords.x1,y=coords.x2), inherit.aes = FALSE, color="red")
#now plot, setting the coordinates to scale correctly and as a black and white plot 
dbplot + theme_bw() + coord_equal()
```
Checking CRS for bradford
```{r}
crs(bradford)
```
Get the bbox in lat long 
```{r}
##First get the bbox in lat long for Bradford
latlong <- "+init=epsg:4326" 
bradfordWGS <-spTransform(bradford, CRS(latlong))
bradfordWGS@bbox
```
Convert basemap to the same CRS
```{r}
basemap<-openmap(c(53.467712,-2.225532),c(53.489362,-2.149995), zoom=NULL,"stamen-toner")
# convert the basemap to British National Grid - remember we created the 
# BNG object right at the beginning of the practical - it's an epsg string...
basemap_bng<-openproj(basemap, projection="+init=epsg:4326")
```
Adding a basemap
```{r}
autoplot(basemap_bng) + geom_point(data=crime_bradford_points, 
                                   aes(coords.x1,coords.x2, 
                                       colour=cluster, fill=cluster)) + 
  geom_polygon(data = chulls, aes(coords.x1,coords.x2, group=cluster, fill=cluster), 
               alpha = 0.5)   + geom_point(data=df_polsub_bradford,aes(x=coords.x1,y=coords.x2), inherit.aes = FALSE, color="red")
```
Plotting all crime occurences in Manchester city
```{r}
tm_shape(manchester_merged_final) +
  tm_polygons(col = NA, alpha = 0.5) +
tm_shape(crime_spdf) +
  tm_dots(col = "blue")
```

# 9. Obtaining Spatial Weights Matrix (entire Manchester city hereafter)

Calculating the area of each polygon (ward) in Manchester city
```{r}
library(rgdal)
getClass("Polygon")
manchester_merged_final$Area_sqkm<-sapply(slot(manchester_merged_final, "polygons"), function(x) sapply(slot(x, "Polygons"), slot, "area"))
str(manchester_merged_final$Area_sqkm)
class(manchester_merged_final$Area_sqkm)
manchester_merged_final$Area_sqkm <- lapply(manchester_merged_final$Area_sqkm, FUN= function(x) x*12100)
```
Extracting the area of each ward data from manchester_merged_final
```{r}
citymanchester.df <- as.data.frame(manchester_merged_final)

citymanchester.df$wardarea <- as.numeric(rownames(citymanchester.df))
class(citymanchester.df$crimes)
  class(citymanchester.df$Area_sqkm)
citymanchester.df$Area_sqkm <- as.numeric(as.character(unlist(citymanchester.df$Area_sqkm)))
citymanchester.df$crimedensity <- citymanchester.df$crimes / citymanchester.df$Area_sqkm
```

Joining crime density data to the manchester_merged_final map
```{r}

```
Reading in the dataset for SES - the average social grade by ward was taken, by using a numeric scale of AB = 1, C1 = 2, C2 = 3, DE = 4, so lower numbers would mean a higher social grade.
```{r}
crimedensity <- read_csv("crimedensity.csv")
summary(crimedensity)
```
Joining crime dennsity dataset with the Manchester basemap 
```{r}
manchester_final <- merge(manchester_merged_final, crimedensity, by.x = "name", by.y = "geography",duplicateGeoms = TRUE)
```

Plotting map to check if third join has succeeded
```{r}
tmap_mode("view")
tm_shape(manchester_final) +
  tm_polygons(col = NA, alpha = 0.5) +
tm_shape(polsub) +
  tm_dots(col = "blue")
```
Adding number of police stations to Manchester_final
```{r}
#reading in data for police stations per ward
library(readr)
policestns <- read_csv("policestns.csv")
manchester_final <- merge(manchester_final, policestns, by.x = "name", by.y = "geography",duplicateGeoms = TRUE)
```

Some clustering in Manchester itself based on KDE diagrams above, and some clustering in Bradford ward (to be analysed later)
Before being able to calculate Moran’s I and any similar statistics, we need to first define a spatial weights matrix
```{r}
library(spdep)
```
Calculating the centroids of all wards in Manchester city
```{r}
coordsWards <- coordinates(manchester_final)
plot(coordsWards)
```

Generating a spatial weights matrix, using contiguity edges corners. 
```{r}
#create a neighbours list
MWard_nb <- poly2nb(manchester_final, queen=T) #this is a neighbours list of queens contiguity
knn_wards <- knearneigh(coordsWards, k=4) #THIS IS A NEIGHBOURS LIST FOR NEAREST NEIGHBOURS CONTIGUITY

MWard_knn <- knn2nb(knn_wards)
#plot neighbours list of queens contiguity
plot(MWard_nb, coordinates(coordsWards), col="red") + plot(manchester_final, add=T)
#plot neighbours list of nearest neighbours contiguity
plot(MWard_knn, coordinates(coordsWards), col="red") + plot(manchester_final, add=T)
```

Create spatial weights object from weights 
```{r}
Mward.queens_weight <- nb2listw(MWard_nb, style="W")
head(Mward.queens_weight)

Mward.nn_weight <- nb2listw(MWard_knn, style="W")
```


# 10. Spatial autocorrelation 
Now we have defined our Wij  matrix, we can calculate the Moran’s I and other associated statistics.

## 10a. Moran's I across Manchester city 
Moran’s I test tells us whether we have clustered values (close to 1) or dispersed values (close to -1), we will calculate for the densities rather than raw values. 
```{r}
I_MWard_Global_Density <- moran.test(manchester_final@data$crimedensity, Mward.queens_weight)
I_MWard_Global_Density
```

## 10b. Geary's C across Manchester city
Geary's C test: insufficient evidence to reject the null hypothesis that similar values of crime densities are clustering
```{r}
C_MWard_Global_Density <- geary.test(manchester_final@data$crimedensity, Mward.queens_weight)
C_MWard_Global_Density
```

Calculate local versions of Moran's I statistic for each ward to see in which wards the hot-spots are
```{r}
#use the localmoran function to generate I for each ward in the city
I_MWard_Local <- localmoran(manchester_final@data$crimes, Mward.queens_weight)
I_MWard_Local_Density <- localmoran(manchester_final@data$crimedensity, Mward.queens_weight)
#what does the output (the localMoran object) look like?
head(I_MWard_Local_Density)
```

Copying I score and p-valule back into the manchester_final spatialpolygonsdataframe
```{r}
manchester_final@data$BLocI <- I_MWard_Local[,1]
manchester_final@data$BLocIz <- I_MWard_Local[,4]
manchester_final@data$BLocIR <- I_MWard_Local_Density[,1]
manchester_final@data$BLocIRz <- I_MWard_Local_Density[,4]
```

Set breaks manually
```{r}
breaks1<-c(-1000,-2.58,-1.96,-1.65,1.65,1.96,2.58,1000)
```

Setting colour palette so higher values are redder
```{r}
MoranColours<- rev(brewer.pal(8, "RdGy"))
```

Plotting the map
```{r}
tm_shape(manchester_final) +
    tm_polygons("BLocIRz",
        style="fixed",
        breaks=breaks1,
        palette=MoranColours,
        midpoint=NA,
        title="Local Moran's I, Crime Occurences in Manchester by ward")
```

# 11. OLS regression

Converting manchester_final to a dataframe
```{r}
manchester_df <- as.data.frame(manchester_final)
#define regression equation so no need to retype each time
reg=crimes~popdens+socialgrade+deprivation+polstns
```
social grade variable's frequency distribution seems to not be normal - negatively skewed (more higher values translating to LOWER social grade). Check the result of a range of transformations along Tukey's ladder.
```{r}
library(car)
symbox(~`socialgrade`, manchester_df, na.rm=T, powers=seq(-3,3,by=.5))
```
deprivation variable's frequency distribution seems to not be normal - positively skewed (more higher values translating to GREATER deprivation). Check the result of a range of transformations along Tukey's ladder.
```{r}
library(car)
symbox(~`deprivation`, manchester_df, na.rm=T, powers=seq(-3,3,by=.5))
```
Trying deprivation^2 for a more normal distribution
```{r}
ggplot(manchester_df, aes(x=(`deprivation`)^2)) + geom_histogram()
```
Plotting it out on a scatterplot
```{r}
qplot(x = (`deprivation`)^2, y = `crimes`, data=manchester_df)
```
qplot(x = (`deprivation`), y = `crimes`, data=manchester_df)
Plotting crimes as dependent variable, OLS for popdens, socialgrade, deprivation
```{r}
m <- lm(log1p(crimes) ~ popdens + deprivation + polstns, data=manchester_df)
summary(m)
```
Check if residuals are normally distributed
```{r}
#save the residuals into your dataframe
manchester_df$m_resids <- m$residuals

qplot(m$residuals) + geom_histogram() 
```
Check for multicollinearity between explanatory variables
```{r}
library(corrplot)
#pull out the columns we want to check for multicolinearity
tempdf <- manchester_df[,c("popdens","deprivation")]


#rename the columns to something shorter
names(tempdf) <- c("Med House Price", "Un Auth Absence")

#compute the correlation matrix for the two variables of interest
cormat <- cor(tempdf[,1:2], use="complete.obs", method="pearson")

#visualise the correlation matrix
corrplot(cormat)
```
Checking VIF to confirm no multicollinearity. 
```{r}
vif(m)
```

# 12. Checking for spatial dependence

## 12a. Checking residuals
Check if any spatial dependence in the residuals, first using queens neighbours
```{r}
lm.morantest(m,Mward.queens_weight)
```
Then using nn
```{r}
lm.morantest(m,Mward.nn_weight)
```

## 12b. Checking using Lagrange Multiplier Tests
Another way to check if necessary to run a spatial model based on the OLS is to do lagrange multiplier tests. 
```{r}
lm.LMtests(m,Mward.nn_weight,test=c("LMerr", "LMlag", "RLMerr", "RLMlag", "SARMA"))
```

# 13. Spatially Lagged X (SLX) model
Start with a spatially lagged x model. Spatially lagged x model – y=Xß+WXT+e.
```{r}
library(spatialreg)
reg2=lmSLX(m,data=manchester_df, Mward.queens_weight)
summary(reg2)
```

# 14. Spatial Lag Model (SLM)
Running the SLM with a queen's case weights matrix. Spatial Lag (Autoregressive) Model - y=pWy+XB+e 
```{r}
reg3=lagsarlm(m,data=manchester_df, Mward.nn_weight)
summary(reg3)

```
Check overall impacts, as infinite feedback loops so p-values for individual variables are meaningless
```{r}
impacts(reg3,listw=Mward.queens_weight)
summary(impacts(reg3,listw=Mward.queens_weight,R=1000),zstats=TRUE) #Add zstats,pvals
```

# 15. Spatial Error Model (SEM)
Next do a SEM
```{r}
reg4=errorsarlm(m,data=manchester_df, Mward.queens_weight)
summary(reg4)
```
# 16. GWR and its plots
Trying GWR
```{r}
library(spgwr)
```
```{r}
#calculate kernel bandwidth
GWRbandwidth <- gwr.sel(log1p(crimes) ~ popdens + deprivation + polstns, data=manchester_df, coords=coordsWards,adapt=T)

```
```{r}
#run the gwr model
gwr.model = gwr(m, coords=coordsWards, adapt=GWRbandwidth, hatmatrix=TRUE, se.fit=TRUE)

#print the results of the model
gwr.model
```

Collect results
```{r}
results<-as.data.frame(gwr.model$SDF)
names(results)
```
Attach results to original dataframe
```{r}
#attach coefficients to original dataframe
manchester_final@data$coefpopdens<-results$popdens
manchester_final@data$coefdeprivation<-results$deprivation
manchester_final@data$coefpolstns<-results$polstns
```

Plotting GWR result for popdens
```{r}
tm_shape(manchester_final) +
  tm_polygons(col = "coefpopdens", palette = "RdBu", alpha = 0.5)
```
Plotting GWR results for deprivation
```{r}
tm_shape(manchester_final) +
  tm_polygons(col = "coefdeprivation", palette = "RdBu", alpha = 0.5) 
```
Plotting GWR results for police stations
```{r}
tm_shape(manchester_final) +
  tm_polygons(col = "coefpolstns", palette = "RdBu", alpha = 0.5) 
```

Calculating standard errors for statistical significance for population density
```{r}
#run the significance test
#sigTest = abs(gwr.model$SDF$"log(`Median House Price (£) - 2014`)") -2 * gwr.model$SDF$"log(`Median House Price (£) - 2014`)_se"

sigTest = abs(gwr.model$SDF$"popdens") -2 * gwr.model$SDF$"popdens_se"

#store significance results
manchester_final$GWRpopdens<-sigTest

```
Plotting significance for population density on the map
```{r}
tm_shape(manchester_final) +
  tm_polygons(col = "GWRpopdens", palette = "RdYlBu")
```
Plotting significance for deprivation on the map
```{r}
#run the significance test
sigTest = abs(gwr.model$SDF$"deprivation") -2 * gwr.model$SDF$"deprivation_se"

#store significance results
manchester_final$GWRdeprivation<-sigTest
#plotting
tm_shape(manchester_final) +
  tm_polygons(col = "GWRdeprivation", palette = "RdYlBu")
sigTest
```
Plotting significance for police stations on the map
```{r}
#run the significance test
sigTest = abs(gwr.model$SDF$"polstns") -2 * gwr.model$SDF$"polstns_se"

#store significance results
manchester_final$GWRpolstns<-sigTest
#plotting
tm_shape(manchester_final) +
  tm_polygons(col = "GWRpolstns", palette = "RdYlBu")
```
